Perspectives on Stewardship

### 2024-08-08 book/Perspectives on Stewardship.tex
draft

>>>>

\section{Understanding Noise by Understanding Inputs}

Noise can be considered in a lot of different ways. However it seems practical to take a leaf out of the study of feedback control in linear time invariant systems \cite{feedback}. At first look these systems seem to be setup quite differently (from a mathematical) perspective than the ones we just went over in the last section. Specifically they are described by a series of ordinary differential equations that are concerned with an input function $u$, an output function $y$, and some parameter (usually time) $t$. One classic example is the dampened spring:

$$m\partial^2y+b\partial y+ky=u$$

This is no longer a direct definition of our transducer but instead a "constraint" governing what our transducer can look like. However given $u$ and the appropriate initial conditions this will give you a single $y(t)$ that describes the spring's path through state space.

As such we can think of $y$ as a specific pathway, determined by $u$, and parametrized by $t$. What is $u$ then? Well it is a force applied to our spring over time and so we can consider it equivalent to setting the input on the transducer from one moment to the next - $u(t)$ is the output of an "input transducer" over time. What's important to notice here however is that $u$ is independent of $y$, i.e. $u$ is specified without knowing what $y$ is.\newline

How does this relate to noise? Well consider the fact that while a die is represented by some kind of distribution that in reality when the dice are rolled we get a specific set of dice rolls. Distributions exist to help us explain phenomena over many repeats of the same "experiment" but they don't really "exist" in real life - we are only ever left with one "pathway" through the state space that those distributions describe. So, if we had some gambling system where the "noise" was generated by dice rolls we can imagine that for any run of the experiment there is a specific $u$ that describes the rolls we got. This $u$ then feeds into our specific gambling system to then give us our path through the state space $y$. If we then want to consider what would happen under another set of dice rolls we'd construct a different $u$ and see what new $y$ we get out of it. But in each case our overall system of $u$ and $y$ stays deterministic and follows all the rules we illuminated in the last section.

What we want to study then is how our system $y$ behaves under different input pathways $u$ and determine which $u$'s lead to bad outcomes. \newline

There is however a problem. In both this case and in the case of building lyupanov functions for the sake of studying stability there is no silver bullet, no single method (or at least no one has found one yet). A great quote from "Feedback Control of Dynamic Systems" is that "the theory only gives the engineer a hunting license to look for a [lyapunov] function" \cite{feedback}. So what are we to do? Well once again we can take a leaf out of the LTI (linear time invariant) systems book. \newline

Their way of dealing with this is to take advantage of the specific properties that make an LTI and LTI - and that's linearity and time invariance. This more or less means that if $u_1$ is one input with response $y_1$ and $u_2$ is an input with response $y_2$ that the response of $au_1 + bu_2$ (where $a$ and $b$ are constants) is given by $ay_1 + by_2$. In other words if a specific input can be represented as the sum of a series of inputs, its response is equal to the sum of the responses of the individual inputs.

So what about the response to the dirac delta function? Suppose we call that $h(t)$. Then because the dirac delta function happens to be an impulse of size 1 that last only infinitesimally long that:

$$u(t) = \int_{-\infty}^{\infty}u(\tau) \delta(\tau)d\tau$$

and therefore:

$$y(t) = \int_{-\infty}^{\infty}u(\tau) h(t-\tau)d\tau$$

More or less all this is saying is that because our system in linear and time invariant that the response to some arbitrary input is the "sum" of the impulse responses multiplied by the magnitude of the input at that time carried forward (or backward) to the current time.

This form in and of itself isn't so interesting but the magic happens when we consider what happens in our input is of the form $e^{st}$.

 $$y(t) = \int_{-\infty}^{\infty}e^{s\tau} h(t-\tau)d\tau$$

 which is equivalent under a change of variable to:

 $$y(t) = \int_{-\infty}^{\infty}e^{s(t-\tau)} h(\tau)d\tau = \int_{-\infty}^{\infty}e^{-s\tau} h(\tau)d\tau e^{st}=H(s)e^{-st}$$

 where $H(s)$ is the Laplace transform of $h(t)$. But notice that we've just shown that for an input $e^{st}$ our system will produce an output that is just some scalar multiple of the input! So what does this $H(s)$ look like? Well we can solve for it in our original equation:

 $$ms^2H(s)e^{st}+bsH(s)e^{st}+kH(s)e^{st}=e^{st}$$

 $$H(s) = \frac{1}{ms^2+bs+k}$$

This $H(s)$ (also known as the transfer function) tells us everything we need to know about how the systems responds to different frequencies of input $s$. For example if $s$ is a root of the above denominator we know we're going to have significant issues. And because we can imagine any arbitrary input $u$ as being some combinations of frequencies we've also sorted out how to tell what properties of the input could give us issues. Very useful indeed.\newline

Notice though that in order to arrive at this result we had to lean heavily on the system being LTI. For some other kind of system none of this would apply. And that's just the nature of studying systems - to find their specific properties there has to be some inherent structure (beyond just having a transducer of some kind) that you can leverage to pull out the characteristics of those systems. \newline

For us then it means that we need to find specific properties of ecosystem subsystems that will allow us to define them as a specific type(s) of system and then leverage those properties into understanding the characteristics of the inputs that could lead to issues. Only by getting specific are we going to be able to make any more progress.
